---
title: "proj3_survival"
date: "2024-03-07"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Goal is to check the analysis of the nejm paper
* And provide additional exploratory analyses that might "break" the results
- Are there other covariates that might lead to a different conclusion
-- PC for dim red on immunology variables
- Are the Kaplan-Meier curves actually different or not
-- Cox pH before 32 weeks and after 34 weeks
- How to deal with the multiple comparison issue
-- Numerous additional analyses on the same dataset
-- Look into different methods of p-value adjustment

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(medicaldata)
library(dplyr)
data('opt')

# colnames(opt)
```

Variables of Interests:
Group
Age: this is age at baseline
Comorbidity indicator: diabetes + hypertension
Drug indicator: drugs addiction + alcohol use
Previos pregnancy
Education

Outcomes:
GA.at.outcome (this is in days): this is gestational age at delivery; use this as event variable
Preg.ended...37.wk: This is the pre-term birth indicator; use this as censoring indicator
- Survival as primary, logistic as secondary

Secondary Outcome:
Birthweight --> need to binarize to less than 2500 g
- Logistic Regression

Check distributions:
Birth.outcome

# Subsetting the Dataframe
```{r echo = FALSE}
serum <- c("OAA1","OCR1","OFN1","OPG1","OPI1","OTD1","OTF1","OCRP1","O1B1","O61","O81","OPGE21","OTNF1","OMMP91","ETXU_CAT1","OFIBRIN1")
opt_copy <- opt[,c('GA.at.outcome', 'Birthweight', 'Age', 'Preg.ended...37.wk', 'Education', 'Hypertension', 'Diabetes', 'Use.Tob', 'Use.Alc', 'Drug.Add', 'Prev.preg', 'Group', serum)]

# Make all spaces and dot (.) NA
opt_copy[opt_copy == ""|opt_copy==" "|opt_copy == "."] <- NA
opt_copy[opt_copy == ''] <- NA


## Change serum variable type to numeric
opt_copy[,serum] <- sapply(opt_copy[,serum], as.numeric)

# Get rid of any trailing spaces
opt_copy <- opt_copy %>% 
  mutate_if(is.factor, as.character) %>% 
  mutate_if(is.character, trimws)

# Re-format Hypertension Variable
opt_copy <- mutate_at(opt_copy, vars(Hypertension), function(x) ifelse(x == "N", "No", ifelse(x == "Y", "Yes", x)))

# Add Comorbidity
# opt_copy$comorbidity <- apply(opt_copy[c('Hypertension', 'Diabetes', 'Prev.preg')], 1, function(x) {ifelse(any(x == "Yes"), "Yes", "No")})
# 
# # Add Drug Indicator
# opt_copy$Drug.Use <- apply(opt_copy[c('Use.Tob', 'Use.Alc', 'Drug.Add')], 1, function(x) {ifelse(any(x == "Yes"), "Yes", "No")})

# Check NAs for each variable
# View(data.frame(apply(opt_copy, 2, function(x){sum(is.na(x))})))
opt_copy[opt_copy == ''] <- NA

# Before dropping NA serum columns, the number of preterm birth is 103
table(opt_copy$Preg.ended...37.wk)

# After dropping NA serum columns, the number of preterm birth is 98
## Dropping missing values in serum varibales
opt_copy_1 <- opt_copy[complete.cases(opt_copy[, serum]), ]
table(opt_copy_1$Preg.ended...37.wk)


complete_preterm <- opt_copy_1 %>% select(-Birthweight) 
complete_preterm <- complete_preterm[complete.cases(complete_preterm),]
complete_preterm <- droplevels(complete_preterm)

complete_birthweight <- opt_copy_1 %>% select(-GA.at.outcome)
complete_birthweight <- complete_birthweight[complete.cases(complete_birthweight),]
complete_birthweight <- droplevels(complete_birthweight)
```

## PCA
```{r}
pca_complete_preterm <- prcomp(complete_preterm[,serum], scale = TRUE)
summary(pca_complete_preterm)
cumvars <- cumsum(pca_complete_preterm$sdev^2 / sum(pca_complete_preterm$sdev^2))

plot(cumvars, xlab = 'Number of PCs', ylab = 'Proportion of Cumulative Variance', main = 'Scree Plot')
grid()
abline(h = 0.8, col = 'blue')

rotated_dat <- as.data.frame(pca_complete_preterm$x)
# 9 PCs instead of 16

complete_preterm_extended <- cbind(complete_preterm[,-c(12:27)], rotated_dat[,1:9])
complete_preterm_extended <- complete_preterm_extended[complete.cases(complete_preterm_extended),]

## For low birth weight
pca_complete_birthweight <- prcomp(complete_birthweight[,serum], scale = TRUE)
summary(pca_complete_birthweight)
cumvars2 <- cumsum(pca_complete_birthweight$sdev^2 / sum(pca_complete_birthweight$sdev^2))

plot(cumvars2, xlab = 'Number of PCs', ylab = 'Proportion of Cumulative Variance', main = 'Scree Plot')
grid()
abline(h = 0.8, col = 'blue')
# 9 PCs instead of 16
rotated_dat <- as.data.frame(pca_complete_birthweight$x)

biplot(pca_complete_birthweight,scale=0,cex=0.5)
abline(h=0,v=0)

complete_birthweight_extended <- cbind(complete_birthweight[,-c(12:27)], rotated_dat[,1:9])
complete_birthweight_extended <- complete_birthweight_extended[complete.cases(complete_birthweight_extended),]

```

# Exploratory Plots for initial data
```{r}
library(survival)
library(survminer)

complete_preterm$Preg.ended...37.wk <- ifelse(complete_preterm$Preg.ended...37.wk == 'Yes', 1, 0)

surv_obj <- with(complete_preterm, Surv(time = GA.at.outcome, event = Preg.ended...37.wk, type = 'right'))

fit <- survfit(surv_obj ~ Group, data = complete_preterm)

ggs <- ggsurvplot(fit,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#F8766D", "#00BFC4"),
           ylim=c(0.825,1),
           xlim=c(90,310),
           title = 'Survival Curves for Control and Treatment Cohort',
           break.x.by = 20)

#ggs$plot <- ggs$plot + ylim(c(0.75,1)) + xlim(c(90, 310))
ggs

surv_diff1 <- survdiff(Surv(time = GA.at.outcome, event = Preg.ended...37.wk, type = 'right') ~ Group, data = complete_preterm)
surv_diff1
```


## Only checking the data pre-32.8 weeks aka 231 days or fewer 
```{r}
pre32 <- complete_preterm

pre32$Preg.ended...37.wk <- NULL
pre32$pre230 <- ifelse(pre32$GA.at.outcome <= 230, 1, 0)

# new survival fit
surv_obj2 <- with(pre32, Surv(time = GA.at.outcome, event = pre230, type = 'right')) 
fit2 <- survfit(surv_obj2 ~ Group, data = pre32)

ggs2 <- ggsurvplot(fit2,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#F8766D", "#00BFC4"),
           ylim=c(0.825,1),
           xlim=c(90,240),
           title = 'Survival Curves for Control and Treatment Cohort',
           break.x.by = 20)

#ggs$plot <- ggs$plot + ylim(c(0.75,1)) + xlim(c(90, 310))
ggs2

surv_diff2 <- survdiff(Surv(time = GA.at.outcome, event = pre230, type = 'right') ~ Group, data = pre32)
surv_diff2
```

## Cox Proportional Hazards for preterm birth

```{r}
fit.coxph <- coxph(Surv(time = GA.at.outcome, event = Preg.ended...37.wk)~., data=complete_preterm)
summary(fit.coxph)

fit.coxph2 <- coxph(Surv(time = GA.at.outcome, event = pre230)~., data=pre32)
# ggsurvplot(survfit(fit.coxph2), data = pre32)
summary(fit.coxph2)
```


## Logistic regression for low birth weight

```{r}
# Define indicator for low birth weight (birth weight less than 2500mg)
complete_birthweight$low.weight = ifelse(complete_birthweight$Birthweight <= 2500, 1, 0)

fit.logis <- glm(low.weight~., data=complete_birthweight[,-1])
summary(fit.logis)
fit.logis.step <- step(fit.logis, direction = "backward")
summary(fit.logis.step)
```



# Exploratory Plots for extended data
```{r}
library(survival)
library(survminer)

complete_preterm_extended$Preg.ended...37.wk <- ifelse(complete_preterm_extended$Preg.ended...37.wk == 'Yes', 1, 0)

surv_obj <- with(complete_preterm_extended, Surv(time = GA.at.outcome, event = Preg.ended...37.wk, type = 'right'))

fit.ext <- survfit(surv_obj ~ Group, data = complete_preterm_extended)

ggs <- ggsurvplot(fit.ext,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#F8766D", "#00BFC4"),
           ylim=c(0.825,1),
           xlim=c(90,310),
           title = 'Survival Curves for Control and Treatment Cohort',
           break.x.by = 20)

#ggs$plot <- ggs$plot + ylim(c(0.75,1)) + xlim(c(90, 310))
ggs

surv_diff1_ext <- survdiff(Surv(time = GA.at.outcome, event = Preg.ended...37.wk, type = 'right') ~ Group, data = complete_preterm_extended)
surv_diff1_ext
```


## Only checking the data pre-32.8 weeks aka 231 days or fewer 
```{r}
pre32_ext <- complete_preterm_extended

pre32_ext$Preg.ended...37.wk <- NULL
pre32_ext$pre230 <- ifelse(pre32_ext$GA.at.outcome <= 230, 1, 0)

# new survival fit
surv_obj2 <- with(pre32_ext, Surv(time = GA.at.outcome, event = pre230, type = 'right')) 
fit2.ext <- survfit(surv_obj2 ~ Group, data = pre32_ext)

ggs2 <- ggsurvplot(fit2.ext,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#F8766D", "#00BFC4"),
           ylim=c(0.825,1),
           xlim=c(90,240),
           title = 'Survival Curves for Control and Treatment Cohort',
           break.x.by = 20)

#ggs$plot <- ggs$plot + ylim(c(0.75,1)) + xlim(c(90, 310))
ggs2

surv_diff2_ext <- survdiff(Surv(time = GA.at.outcome, event = pre230, type = 'right') ~ Group, data = pre32_ext)
surv_diff2_ext
```

## Cox Proportional Hazards for preterm birth

```{r}
# Full-length of the observation period
fit.coxph.extend <- coxph(Surv(time = GA.at.outcome, event = Preg.ended...37.wk)~., data=complete_preterm_extended)
summary(fit.coxph.extend)
coef = round(cbind(summary(fit.coxph.extend)$coefficients[, c(2,5)], summary(fit.coxph.extend)$conf.int[, c(3,4)]),3)
colnames(coef) = c("Hazard","P value", "2.5%", "97.5%")
knitr::kable(coef)

# 33 weeks of the observation period
fit.coxph2.extend <- coxph(Surv(time = GA.at.outcome, event = pre230)~., data=pre32_ext)
summary(fit.coxph2.extend)
coef2 = round(cbind(summary(fit.coxph2.extend)$coefficients[, c(2,5)], summary(fit.coxph2.extend)$conf.int[, c(3,4)]),3)
colnames(coef2) = c("Hazard","P value", "2.5%", "97.5%")
knitr::kable(coef2)
```


## Logistic regression for low birth weight

```{r}
# Define indicator for low birth weight (birth weight less than 2500mg)
complete_birthweight_extended$low.weight = ifelse(complete_birthweight_extended$Birthweight <= 2500, 1, 0)

fit.logis.extend <- glm(low.weight~., data=complete_birthweight_extended[,-1])
summary(fit.logis.extend)
fit.logis.step.extend <- step(fit.logis.extend, direction = "backward")
summary(fit.logis.step.extend)

coef.log = round(cbind(exp(summary(fit.logis.step.extend)$coefficients[,1]),drop1(fit.logis.step.extend, test = "Chisq")$`Pr(>Chi)`, exp(confint(fit.logis.step.extend))),3)
colnames(coef.log) = c("Odds ratio","P value", "2.5%", "97.5%")
knitr::kable(coef.log)
```



# Sensitivity analysis

```{r, message=FALSE}
# Check missing value for the initial data
# View(data.frame(apply(opt_copy, 2, function(x){sum(is.na(x))})))

# Multiple imputation
library(mice)

opt_copy <- opt_copy %>% mutate_if(is.character, as.factor)

my_imp = mice(opt_copy, m=5, maxit = 10, seed = 2024)
opt_imp = complete(my_imp, 1)

# Check missing value for imputed data
# View(data.frame(apply(opt_imp, 2, function(x){sum(is.na(x))})))


itt_preterm <- opt_imp %>% select(-Birthweight)
itt_birthweight <- opt_imp %>% select(-GA.at.outcome)
```


## PCA
```{r}
pca_itt_preterm <- prcomp(itt_preterm[,serum], scale = TRUE)
summary(pca_itt_preterm)
cumvars <- cumsum(pca_itt_preterm$sdev^2 / sum(pca_itt_preterm$sdev^2))

plot(cumvars, xlab = 'Number of PCs', ylab = 'Proportion of Cumulative Variance', main = 'Scree Plot')
grid()
abline(h = 0.8, col = 'blue')

rotated_dat <- as.data.frame(pca_itt_preterm$x)
# 9 PCs instead of 16

itt_preterm_extended <- cbind(itt_preterm[,-c(12:27)], rotated_dat[,1:9])
itt_preterm_extended <- itt_preterm_extended[complete.cases(itt_preterm_extended),]

## For low birth weight
pca_itt_birthweight <- prcomp(itt_birthweight[,serum], scale = TRUE)
summary(pca_itt_birthweight)
cumvars2 <- cumsum(pca_itt_birthweight$sdev^2 / sum(pca_itt_birthweight$sdev^2))

plot(cumvars2, xlab = 'Number of PCs', ylab = 'Proportion of Cumulative Variance', main = 'Scree Plot')
grid()
abline(h = 0.8, col = 'blue')
# 9 PCs instead of 16
rotated_dat <- as.data.frame(pca_itt_birthweight$x)

itt_birthweight_extended <- cbind(itt_birthweight[,-c(12:27)], rotated_dat[,1:9])
itt_birthweight_extended <- itt_birthweight_extended[complete.cases(itt_birthweight_extended),]
```

## Logrank and plots

```{r}
itt_preterm_extended$Preg.ended...37.wk <- ifelse(itt_preterm_extended$Preg.ended...37.wk == 'Yes', 1, 0)

surv_obj <- with(itt_preterm_extended, Surv(time = GA.at.outcome, event = Preg.ended...37.wk, type = 'right'))

fit.ext <- survfit(surv_obj ~ Group, data = itt_preterm_extended)

ggs <- ggsurvplot(fit.ext,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#F8766D", "#00BFC4"),
           ylim=c(0.825,1),
           xlim=c(90,310),
           title = 'Survival Curves for Control and Treatment Cohort',
           break.x.by = 20)

#ggs$plot <- ggs$plot + ylim(c(0.75,1)) + xlim(c(90, 310))
ggs

surv_diff1_ext <- survdiff(Surv(time = GA.at.outcome, event = Preg.ended...37.wk, type = 'right') ~ Group, data = itt_preterm_extended)
surv_diff1_ext
```


## Only checking the data pre-32.8 weeks aka 231 days or fewer 
```{r}
pre32_ext <- itt_preterm_extended

pre32_ext$Preg.ended...37.wk <- NULL
pre32_ext$pre230 <- ifelse(pre32_ext$GA.at.outcome <= 230, 1, 0)

# new survival fit
surv_obj2 <- with(pre32_ext, Surv(time = GA.at.outcome, event = pre230, type = 'right')) 
fit2.ext <- survfit(surv_obj2 ~ Group, data = pre32_ext)

ggs2 <- ggsurvplot(fit2.ext,
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#F8766D", "#00BFC4"),
           ylim=c(0.825,1),
           xlim=c(90,240),
           title = 'Survival Curves for Control and Treatment Cohort',
           break.x.by = 20)

#ggs$plot <- ggs$plot + ylim(c(0.75,1)) + xlim(c(90, 310))
ggs2

surv_diff2_ext <- survdiff(Surv(time = GA.at.outcome, event = pre230, type = 'right') ~ Group, data = pre32_ext)
surv_diff2_ext
```

## Cox Proportional Hazards for preterm birth

```{r}
fit.coxph.extend <- coxph(Surv(time = GA.at.outcome, event = Preg.ended...37.wk)~., data=itt_preterm_extended)
summary(fit.coxph.extend)
coef = round(cbind(summary(fit.coxph.extend)$coefficients[, c(2,5)], summary(fit.coxph.extend)$conf.int[, c(3,4)]),3)
colnames(coef) = c("Hazard","P value", "2.5%", "97.5%")
knitr::kable(coef)

fit.coxph2.extend <- coxph(Surv(time = GA.at.outcome, event = pre230)~., data=pre32_ext)
summary(fit.coxph2.extend)
coef2 = round(cbind(summary(fit.coxph2.extend)$coefficients[, c(2,5)], summary(fit.coxph2.extend)$conf.int[, c(3,4)]),3)
colnames(coef2) = c("Hazard","P value", "2.5%", "97.5%")
knitr::kable(coef2)
```


## Logistic regression for low birth weight

```{r}
# Define indicator for low birth weight (birth weight less than 2500mg)
itt_birthweight_extended$low.weight = ifelse(itt_birthweight_extended$Birthweight <= 2500, 1, 0)

# fit.logis.extend <- glm(low.weight~., data=itt_birthweight_extended[,-1])
# summary(fit.logis.extend)
# fit.logis.step.extend <- step(fit.logis.extend, direction = "backward")
# summary(fit.logis.step.extend)


fit.logis.itt <- glm(low.weight~Preg.ended...37.wk+Diabetes+PC3+PC7+PC8, data=itt_birthweight_extended[,-1])
summary(fit.logis.itt)

coef.log2 = round(cbind(exp(summary(fit.logis.itt)$coefficients[,1]),drop1(fit.logis.itt, test = "Chisq")$`Pr(>Chi)`, exp(confint(fit.logis.itt))),3)
colnames(coef.log2) = c("Odds ratio","P value", "2.5%", "97.5%")
knitr::kable(coef.log2)
```


